Lazy Linearity for a Core Functional Language

### 2025-10-17 popl26.tex
more tweaks
More on cases

In Core, cases are of the form $\ccase{e_s}{z~\{\ov{\rho_i \to e_i}\}}$,
where $e_s$ is the case \emph{scrutinee}, $z$ is the case \emph{binder}, and
$\ov{\rho_i \to e_i}$ are the case \emph{alternatives}, composed of a pattern
$\rho_i$ and of the corresponding expression $e_i$. 
%\begin{enumerate}
%\item
The case scrutinee is always evaluated to Weak Head Normal Form (WHNF) and
%
% \item
% evaluating an expression that is already in WHNF is a no-op, that is,
% no computation occurs; 
%
the case binder is an alias to the result of evaluating the scrutinee
to WHNF. Moreover, in Core, the alternative patterns are always exhaustive.
% i.e. there always exists
% a pattern that matches the WHNF of a value resulting from evaluating the
% scrutinee, where a pattern is either a wildcard that matches all expressions
% ($\_$), or a constructor and its linear and non-linear component binders
% ($K~\ov{x}\ov{y}$, with $\ov{x}$ as linearly-bound variables and $\ov{y}$ as
% unrestricted ones).
%
%\end{enumerate}
%
% ROMES:TODO: We should talk about WHNF in the background, not here.
% \parawith{WHNF} An expression is in Weak Head Normal Form when 
%
\ROUNDTWO{The first} example program constructs a tuple from linear resources,
matches on it, then uses both linearly-bound variables from the tuple pattern.
This is well-typed in Linear Haskell:
\begin{working}
\begin{lstlisting}
f6 :: a *' $\lolli$ '* b *' $\lolli$ '* (a *' $\lolli$ '* b *' $\lolli$ '* c) -> c
f6 x y use = case (x,y) of z { (a,b) -> use a b }
\end{lstlisting}
\end{working}
\ROUNDTWO{Perhaps surprisingly}, a similar program which discards the pattern
variables, and instead uses the resources \ROUNDTWO{from} the scrutinee
\ROUNDTWO{again}, is still linear, despite not being accepted by Linear Haskell:
\begin{notyet}
\begin{lstlisting}
f7 :: a *' $\lolli$ '* b *' $\lolli$ '* (a *' $\lolli$ '* b *' $\lolli$ '* c) -> c
f7 x y use = case (x,y) of z { (a,b) -> use x y }
\end{lstlisting}
\end{notyet}
\incode{f7} \ROUNDTWO{can be seen as} linear by \ROUNDTWO{realizing} that
the tuple being scrutinized is already in WHNF, so evaluating it again does
not consume either $x$ \ROUNDTWO{nor} $y$. Even if the tuple \ROUNDTWO{paired}
two \ROUNDTWO{arbitrary} expressions which \ROUNDTWO{captured} $x$ and $y$,
\ROUNDTWO{because $a$ and $b$ are not forced in the continuation, $x$ and $y$ would remain unused.}
\ROUNDTWO{On the other hand, as seen in the next example}, if $a$ were used in 
the case body, then $x$ \ROUNDTWO{is consumed via $a$ and must not be used again.}
Thus, \incode{f8} must be rejected:
\begin{noway}
\begin{lstlisting}
f8 :: a *' $\lolli$ '* b *' $\lolli$ '* (a *' $\lolli$ '* a *' $\lolli$ '* c) -> c
f8 x y use = case (x,y) of z { (a,b) -> use a x }
\end{lstlisting}
\end{noway}
This idea that $x$ and $a$ are mutually exclusive is \ROUNDTWO{akin to how} let
bindings \ROUNDTWO{are} mutually exclusive \ROUNDTWO{with the set} of resources
\ROUNDTWO{they capture, in the let-continuation}.
\ROUNDTWO{Forcing a pattern variable evaluates the corresponding expression to
which the constructor was applied, and all linear variables captured by that
expression may }.
Otherwise, if we do not use the
pattern variables, then we do not run the computation and thus no resources are
consumed.

A third (semantically well-typed, but rejected by Linear Haskell) option is to use the case binder $z$ instead of $a,b$ or $x,y$:
\begin{notyet}
\begin{lstlisting}
f9 :: a *' $\lolli$ '* b *' $\lolli$ '* (a *' $\lolli$ '* b *' $\lolli$ '* c) -> c
f9 x y use = case (x,y) of z { (a,b) -> uncurry use z }
\end{lstlisting}
\end{notyet}
Again, $z$ is mutually exclusive with $(a,b)$ and with $(x,y)$, but at least one of
the three must occur to ensure the linear resources are consumed. In this
example, we can think that using $a$ entails using the resource $x$, $b$ the
resource $y$, and the case binder $z$ entails using both $a$ and $b$.

@@

\subsubsection{Branching on WHNF-ness}\label{sec:onwhnf}

>>>>

In Core, cases are of the form $\ccase{e_s}{z~\{\ov{\rho_i \to e_i}\}}$,
where $e_s$ is the case \emph{scrutinee}, $z$ is the case \emph{binder}, and
$\ov{\rho_i \to e_i}$ are the case \emph{alternatives}, composed of a pattern
$\rho_i$ and of the corresponding expression $e_i$. 
%\begin{enumerate}
%\item
The case scrutinee is always evaluated to Weak Head Normal Form (WHNF) and
%
% \item
% evaluating an expression that is already in WHNF is a no-op, that is,
% no computation occurs; 
%
the case binder is an alias to the result of evaluating the scrutinee
to WHNF. Moreover, in Core, the alternative patterns are always exhaustive.
% i.e. there always exists
% a pattern that matches the WHNF of a value resulting from evaluating the
% scrutinee, where a pattern is either a wildcard that matches all expressions
% ($\_$), or a constructor and its linear and non-linear component binders
% ($K~\ov{x}\ov{y}$, with $\ov{x}$ as linearly-bound variables and $\ov{y}$ as
% unrestricted ones).
%
%\end{enumerate}
%
% ROMES:TODO: We should talk about WHNF in the background, not here.
% \parawith{WHNF} An expression is in Weak Head Normal Form when 
%
\ROUNDTWO{The first} example program constructs a tuple from linear resources,
matches on it, then uses both linearly-bound variables from the tuple pattern.
This is well-typed in Linear Haskell:
\begin{working}
\begin{lstlisting}
f6 :: a *' $\lolli$ '* b *' $\lolli$ '* (a *' $\lolli$ '* b *' $\lolli$ '* c) -> c
f6 x y use = case (x,y) of z { (a,b) -> use a b }
\end{lstlisting}
\end{working}
\ROUNDTWO{Perhaps surprisingly}, a similar program which discards the pattern
variables, and instead uses the resources \ROUNDTWO{from} the scrutinee
\ROUNDTWO{again}, is still linear, despite not being accepted by Linear Haskell:
\begin{notyet}
\begin{lstlisting}
f7 :: a *' $\lolli$ '* b *' $\lolli$ '* (a *' $\lolli$ '* b *' $\lolli$ '* c) -> c
f7 x y use = case (x,y) of z { (a,b) -> use x y }
\end{lstlisting}
\end{notyet}
\incode{f7} \ROUNDTWO{can be seen as} linear by \ROUNDTWO{realizing} that
the tuple being scrutinized is already in WHNF, so evaluating it again does
not consume either $x$ \ROUNDTWO{nor} $y$. Even if the tuple \ROUNDTWO{paired}
two \ROUNDTWO{arbitrary} expressions which \ROUNDTWO{captured} $x$ and $y$,
\ROUNDTWO{because $a$ and $b$ are not forced in the continuation, $x$ and $y$ would remain unused.}
\ROUNDTWO{On the other hand, as seen in the next example}, if $a$ were used in 
the case body, then $x$ \ROUNDTWO{is consumed via $a$ and must not be used again.}
Thus, \incode{f8} must be rejected:
\begin{noway}
\begin{lstlisting}
f8 :: a *' $\lolli$ '* b *' $\lolli$ '* (a *' $\lolli$ '* a *' $\lolli$ '* c) -> c
f8 x y use = case (x,y) of z { (a,b) -> use a x }
\end{lstlisting}
\end{noway}
This idea that $x$ and $a$ are mutually exclusive is \ROUNDTWO{akin to how} let
bindings \ROUNDTWO{are} mutually exclusive \ROUNDTWO{with the set} of resources
\ROUNDTWO{they capture, in the let-continuation}.
\ROUNDTWO{Forcing a pattern variable evaluates the expression for the
corresponding constructor field, and all linear variables captured by that
expression may end up being used}.
% If we do not use the pattern variables, then we do not run the computation and
% thus no resources are consumed.
%
A third (semantically well-typed, but rejected by Linear Haskell) option is to use the case binder $z$ instead of $a,b$ or $x,y$:
\begin{notyet}
\begin{lstlisting}
f9 :: a *' $\lolli$ '* b *' $\lolli$ '* (a *' $\lolli$ '* b *' $\lolli$ '* c) -> c
f9 x y use = case (x,y) of z { (a,b) -> uncurry use z }
\end{lstlisting}
\end{notyet}
Again, $z$ is mutually exclusive with $(a,b)$ and with $(x,y)$, but at least
one of the three must \ROUNDTWO{be used} to ensure the
linear resources are consumed. Essentially, in this example, using $a$ entails
using the resource $x$, $b$ the resource $y$, and the case binder $z$ entails
using both $a$ and $b$.

@@

\subsubsection{\ROUNDTWO{Discriminating by} WHNF-ness}\label{sec:onwhnf}
